---
title: "Exam 2"
author: "Nate Hawkins"
date: "3/14/2021"
output:
  word_document: default
  pdf_document: default
---

```{r}
library(R2jags)
library(dplyr)
library(brms)
library(loo)
library(rstan)
```


This is a take-home exam.  The exam is due on Tuesday, March 23, at 9:30 am.  Please complete the exam
using a Markdown file.  Please change the **author** field in the .Rmd file from "Stat 451" to your name.
Please email your 
completed exam to gwf@byu.edu and to dteuscher.37.12@gmail.com prior to the 9:30 am deadline.  Please email both your Mardkown file (.Rmd) and
your output file.  You may use whatever format is most convenient for your output document.  
Use the following convention to name your files (assumeing you chose .html as your output format):
'lastname_exam1.Rmd' and 'lastname_exam1.html'
where lastname is replaced with your last name.  Make sure you show your code as well as your answers.

Take-home exams should
be your own work.  However, you are welcome to use class notes, and help documentation publicly 
available for all the programs we have used.  You should not search the web for similar problems
which someone else may have solved.  You should not discuss the exam with any living person.  
The data for the exam are in data files that I will email to you.

You are responsible for making sure that your chains have converged appropriately by examining such things as
the Raftery-Louis diagnostic and the effective sample size.  You will not be asked to do this formally in the 
exam, but if your chains are not correct, you will get wrong answers.

For the first set of questions we will be using the data file **influent.dat**.  You should have received that
data file with the exam.  Water flows into the Mississippi river from a number of streams and rivers. These
sources carry nitrogen into the river.  In the data set, there are six sources of the nitrogen, which 
are a random sample of the many hundreds of streams and rivers that flow into the Mississippi.  These 
sources have been classified by type.  The three types are: (1) no farm land in watershed, (2) less than
50% farm land in watershed, and (3) more than 50% farmland in watershed.  These three types are the only
types that we seek to analyze.  In the data file, the three columns are river source, nitrogen, and type.

1. First read in the data and print out the first six rows of the data.

```{r}
influent <- read.table("influent.dat")
colnames(influent) <- c("source", "nitrogen", "type")

```

#Need to check this without adapt_delta

2.  Ignoring the source, use brm to write a model to find the differences in type.  Use priors of
normal(0,100) for the type effects, and gamma(2,.1) for $\sigma_{error}$.  Remember that the kind of
variable you are working with will make a difference. What is the level of nitrogen estimated for type 3.

Level of nitrogen estimated for type 3 is 36.36

```{r}
source = influent$source
nitrogen = influent$nitrogen
type = influent$type
influent$vtype = as.factor(influent$type)


fit1 <- brm(formula = nitrogen ~ -1 + vtype, data = influent, 
            family = "gaussian",
            prior = c(set_prior("normal(0,100)", class = "b"),
                      set_prior("gamma(2,.1)", class = "sigma")),
            warmup = 1000, iter = 5000, chains = 4, 
            #control = list(adapt_delta = 0.98),
            save_pars = save_pars(all = TRUE))

summary(fit1)

```

```{r, eval = FALSE}
chains <- as.matrix(fit1)
dim(chains)
head(chains)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)
```


3. What is the looic for this model?

The looic is 255.6
```{r}
loo1 <- loo(fit1)
print(loo1)
```


4.  Now run the same model using JAGS.  Use the same priors for the type parameters (remember,
normal priors in JAGS use precision), but use a gamma(5,.1) for $\sigma^2_{error}$.  What is the level of nitrogen estimated for type 3?

```{r}
mdl <- "

model {

  for(i in 1:37){
    nitrogen[i] ~ dnorm(mu[i], 1/s2error)
    mu[i] <- beta[type[i]]
  }
  
  # Priors
  for(i in 1:3){
    beta[i] ~ dnorm(0, 0.0001)
  }
  
  s2error ~ dgamma(5, 0.1)

}
"

writeLines(mdl, 'fit2.txt')

source = influent$source
nitrogen = influent$nitrogen
type = influent$type

data.jags <- c('nitrogen', 'type')
parms <- c('beta' ,'s2error')

fit2 <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'fit2.txt', inits = NULL,
                  n.iter = 20000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
fit2
```

```{r eval = FALSE}
sims <- as.mcmc(fit2)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)
```


5.  What is the DIC of the above model?

DIC is 255.3

6.  Now we want to account for the variance in sources to make inference.  Redo the model in 
brm, but now put source in the model appropriately.  Use a gamma(2,.1) prior for $\sigma_{source}$.
The other priors can stay the same.  Now what is the estimate for the type 3 mean?

Estimate for type 3 mean is 12.24.
This plus whatever the effect for whatever source it is.

```{r}

fit3 <- brm(formula = nitrogen ~ -1 + vtype + (1|source), data = influent, 
            family = "gaussian",
            prior = c(set_prior("normal(0,10)", class = "b"),
                      set_prior("gamma(2,.1)", class = "sigma"),
                      set_prior("gamma(2,.1)", class = "sd")),
            warmup = 1000, iter = 20000, chains = 4, thin = 10,
            control = list(adapt_delta = 0.98),
            save_pars = save_pars(all = TRUE))
summary(fit3)

fit3$fit
```


```{r}
# Diagnostics look good
chains <- as.matrix(fit3)
dim(chains)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)

```


7.  What is the looic for this model?

looic is 252.7

```{r}
loo3 <- loo(fit3)
```


8.  Which model (problem 2 or problem 6 is better?  Why?

The model for problem 6 is better. It has a looic of 252.7 compared to 255.3. Clearly accounting for source is important.

9.  Now do the model with the variance for sources accounted for in JAGS.  Use
a gamma(5,.1) for the variance component $\sigma^2_{source}$. What is the 
estimate for the type 3 mean?

Estimate for type 3 mean is 36.24.

```{r}
mdl <- "

model {

  for(i in 1:37){
    nitrogen[i] ~ dnorm(mu[i], 1/s2error)
    mu[i] <- beta[type[i]] + u[source[i]]
  }
  
  # Priors
  for(i in 1:3){
    beta[i] ~ dnorm(0, 0.0001)
  }
  for(i in 1:6){
    u[i] ~ dnorm(0,1/s_source)
  }
  
  s2error ~ dgamma(5, 0.1)
  s_source ~ dgamma(5,0.1)

}
"

writeLines(mdl, 'fit4.txt')

source = influent$source
nitrogen = influent$nitrogen
type = influent$type

data.jags <- c('nitrogen', 'type', 'source')
parms <- c('beta' , 'u', 's2error')

fit4 <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'fit4.txt', inits = NULL,
                  n.iter = 20000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
(fit4)
```

```{r, eval = FALSE}
sims <- as.mcmc(fit4)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)
```



10. What is the DIC for this model?

DIC is 252.0

11.  Which model (problem 4 or problem 9) is better?  Why?

DIC for model 9 is better, it is lower. Therefore there is value in accounting for the which source it comes from. We can now make inference for sources.

We will be using the data file **pig.dat** for the the next set of questions.  You should have received that data file.
The data set is from a feeding experiment comparing 3 treatments (basically how much of a particular additive was included
in the food ration) on young hogs over the first 30 days of life.  The measurements were taken every 6 days.  The pig number
identifies the pig within a treatment.  That is, pig 1 in trt 1 is not the same pig as pig 1 in trt 2.

12.  Plot the weight on the y-axis and the day  on the x-axis, ignoring the trt and the pig. 

```{r}
pig <- read.table("pig.dat", header = TRUE)
head(pig)
ggplot(data = pig, mapping  = aes(x = day, y = weight)) + 
  geom_point()
```


13.  You will note from your plot that weight gain is fairly linear over the first 15 days or so, and then tails off.
To account for this, we are going to add a quadratic term for day.  That is, include a column for day^2 in the data
set. Print the first 6 rows of the new data set.

```{r}
pig$day2 = (pig$day)^2
head(pig)
```


14. Using JAGS, run a model with an intercept, linear term, and quadratic term for the overall data.  You are essentially
fitting the data you plotted, with no concern about treatments or the multiple observations on each pig.  Use dnorm(0,0.0001) priors
for the coefficients of the model
(the $\beta$'s), and a dgamma(1.1,.1) as the prior for $\sigma^2_{error}$.  Print a summary of your model.  What is the
DIC of the model?  

DIC is 2073

```{r}
mdl <- "
  model {
  
  for (i in 1:360){
    weight[i] ~ dnorm(mu[i], 1/vv)
    mu[i] <- b0 + b1*day[i] + b2*day2[i]
  }
  
  b0 ~ dnorm(0, 0.0001)
  b1 ~ dnorm(0, 0.0001)
  b2 ~ dnorm(0, 0.0001)
  vv ~ dgamma(1.1, 0.1)
  }

"
day = pig$day
day2 = pig$day2
weight = pig$weight

writeLines(mdl, 'fit5.txt')


data.jags <- c('weight', 'day', 'day2')
parms <- c('b0' , 'b1', 'b2', 'vv')

fit5 <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'fit5.txt', inits = NULL,
                  n.iter = 12000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
(fit5)
```

```{r, eval = FALSE}
sims <- as.mcmc(fit5)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)

```


15.  Run the same model using brm.  Use the same priors for the $\beta$'s (which will now be normal(0,100) since brm
works in standard deviations), and a gamma(1.1,.5) prior for the 
standard deviation ($sigma_{error}$).  Print a summary of your model. What is the looic of the model?

looic is 2073.6

```{r}

fit6 <- brm(formula = weight ~  day + day2, data = pig, 
            family = "gaussian",
            prior = c(set_prior("normal(0,100)", class = "b"),
                      #set_prior("normal(0,100)", class = "a"),
                      set_prior("gamma(1.1,0.5)", class = "sigma")),
            warmup = 1000, iter = 5000, chains = 4, 
            #control = list(adapt_delta = 0.98),
            save_pars = save_pars(all = TRUE), silent = TRUE)

(fit6)

```

```{r, eval = FALSE}

# Convergence and ESS are good
chains <- as.matrix(fit6)
dim(chains)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)

# looic
loo(fit6)
```


16. Using the estimated coefficients from either the JAGS or the brm model, add a best fit line to the plot you
made in problem 12.

```{r}

ggplot(data = pig, mapping  = aes(x = day, y = weight)) + 
  geom_point() + 
  geom_abline(intercept = 14.73, slope = 1.69)



```


17.  Using JAGS, run a model with an intercept, linear slope, and quadratic slope for each of the trt's.  That is,
you will be computing 9 $\beta$'s, 3 for each trt.  Assume all the data are independent (that is, we are not
worried about the different pigs at this point).  Use dnorm(0,0.0001) priors for the coefficients of the model
(the $\beta$'s), and a dgamma(1.1,.1) as the prior for $\sigma^2_{error}$.  While you have now
accounted for the different treatments, you are still ignoring that there are multiple measurements on each animal.  
What is the DIC for this model?

DIC is 1977.4

```{r}
mdl <- "
  model {
  
  for (i in 1:360){
    weight[i] ~ dnorm(mu[i], 1/vv)
    mu[i] <- b0[trt[i]] + b1[trt[i]]*day[i] + b2[trt[i]]*day2[i]
  }
  
  for(i in 1:3){
    b0[i] ~ dnorm(0, 0.0001)
    b1[i] ~ dnorm(0, 0.0001)
    b2[i] ~ dnorm(0, 0.0001)
  }
  
  vv ~ dgamma(1.1, 0.1)
  }

"
trt = pig$trt
day = pig$day
day2 = pig$day2
weight = pig$weight

writeLines(mdl, 'fit7.txt')


data.jags <- c('weight', 'day', 'day2', 'trt')
parms <- c('b0' , 'b1', 'b2', 'vv')

fit7 <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'fit7.txt', inits = NULL,
                  n.iter = 12000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
fit7

```

```{r, eval = FALSE}

# Diagnostics are good
sims <- as.mcmc(fit7)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)
```

18. Repeat this analysis using brm.  You will need to create a factor variable from trt (that is, let a variable be as.factor(pig$trt)).
Use the same priors for the $\beta$'s and a gamma(1.1,.5) as the prior for $\sigma^2_{error}$.  Print a summary of the model.  What is the looic?

looic is 1978.6
```{r}
pig$trtf = as.factor(pig$trt)
fit8 <- brm(formula = weight ~  -1 + trtf + trtf:day + trtf:day2 , data = pig, 
            family = "gaussian",
            prior = c(set_prior("normal(0,100)", class = "b"),
                      set_prior("gamma(1.1,0.5)", class = "sigma")),
            warmup = 1000, iter = 5000, chains = 4, 
            #control = list(adapt_delta = 0.98),
            save_pars = save_pars(all = TRUE))
summary(fit8)

```

```{r, eval = FALSE}

# Convergence and ESS are good
chains <- as.matrix(fit8)
dim(chains)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)

```

```{r}
# looic
loo(fit8)
```


19.  Now we are going to take into account that there are multiple measures on each animal to properly account for this information
when we make inference.  Using JAGS, create a term for a random deviation from the intercept for the intercepts only (this is sometimes
referred to as a random coefficients approach).  That is, we are assuming each pig is a random
draw from the population of all pigs.  You will need to account for the numbering of the pigs in the data set, since 
pig 1 in trt1 is not the same as pig 1 in trt 2.  In this model you will be estimating 60 intercepts (20 pigs in each of the
three treatments).  Use a gamma(1.1,.1)  prior for $\sigma^2_{intercepts}$.  You will also need a prior for all the
terms that are deviations from the overall intercept.  Use a normal with a mean of 0, and a precision that is $1/\sigma^2_{intercepts}$.
What is the DIC?

DIC is 1696.2

```{r}

pig$newpig <- rep(1:60, each = 6)


mdl <- "
  model {
  
  for (i in 1:360){
    weight[i] ~ dnorm(mu[i], 1/vv)
    mu[i] <- b0[trt[i]] + b1[trt[i]]*day[i] + b2[trt[i]]*day2[i] + u0[newpig[i]]
  }
  
  for(i in 1:3){
    b0[i] ~ dnorm(0, 0.0001)
    b1[i] ~ dnorm(0, 0.0001)
    b2[i] ~ dnorm(0, 0.0001)
  }
  
  for(i in 1:60){
    u0[i] ~ dnorm(0,1/vvint)
  }
  
  vvint ~ dgamma(1.1, 0.1)
  vv ~ dgamma(1.1, 0.1)
}

"
newpig <- pig$newpig
trt = pig$trt
day = pig$day
day2 = pig$day2
weight = pig$weight

writeLines(mdl, 'fit9.txt')


data.jags <- c('weight', 'day', 'day2', 'trt', 'newpig')
parms <- c('b0' , 'b1', 'b2','u0', 'vv', 'vvint')

fit9 <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'fit9.txt', inits = NULL,
                  n.iter = 8000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
fit9

```

```{r, eval = FALSE}

# Diagnostics are good
sims <- as.mcmc(fit9)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)
```


20. Continuing in JAGS, add a random coefficient term for the deviations from the linear term in the model.  You will now
have random coefficients for both the intercepts and the linear term (slopes).  You can use a gamma(1.1,.1)  prior for $\sigma^2_{linear}$.
You will also need a prior for all the
terms that are deviations from the overall linear term (slopes) .  Use a normal with a mean of 0, and a precision that is $1/\sigma^2_{linear}$. 
What is the DIC?

DIC is 1451.0

```{r}
mdl <- "
  model {
  
  for (i in 1:360){
    weight[i] ~ dnorm(mu[i], 1/vv)
    mu[i] <- b0[trt[i]] + b1[trt[i]]*day[i] + b2[trt[i]]*day2[i] + u0[newpig[i]] + u1[newpig[i]]*day[i]
  }
  
  for(i in 1:3){
    b0[i] ~ dnorm(0, 0.0001)
    b1[i] ~ dnorm(0, 0.0001)
    b2[i] ~ dnorm(0, 0.0001)
  }
  
  for(i in 1:60){
    u0[i] ~ dnorm(0,1/vvint)
    u1[i] ~ dnorm(0, 1/vvslp)
  }
  
  vvslp ~ dgamma(1.1, 0.1)
  vvint ~ dgamma(1.1, 0.1)
  vv ~ dgamma(1.1, 0.1)
}

"
newpig <- pig$newpig
trt = pig$trt
day = pig$day
day2 = pig$day2
weight = pig$weight

writeLines(mdl, 'fit10.txt')


data.jags <- c('weight', 'day', 'day2', 'trt', 'newpig')
parms <- c('b0' , 'b1', 'b2','u0','u1', 'vv', 'vvint', 'vvslp')

fit10 <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'fit10.txt', inits = NULL,
                  n.iter = 20000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
fit10

```

```{r, eval = FALSE}

# Diagnostics are good
sims <- as.mcmc(fit10)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)
```


21.  Continuing in JAGS, add a random coefficient term for the deviations from the quadratic term in the model.  You will now
have random coefficients for the intercepts, the linear term (slopes), and the quadratic term.  You can use a gamma(1.1,.1)  prior for $\sigma^2_{quadratic}$.
You will also need a prior for all the
terms that are deviations from the overall quadratic term.  Use a normal with a mean of 0, and a precision that is $1/\sigma^2_{quadratic}$. 
What is the DIC?

DIC is 1476.5

```{r}
mdl <- "
  model {
  
  for (i in 1:360){
    weight[i] ~ dnorm(mu[i], 1/vv)
    mu[i] <- b0[trt[i]] + b1[trt[i]]*day[i] + b2[trt[i]]*day2[i] + u0[newpig[i]] + u1[newpig[i]]*day[i] + u2[newpig[i]]*day2[i]
  }
  
  for(i in 1:3){
    b0[i] ~ dnorm(0, 0.0001)
    b1[i] ~ dnorm(0, 0.0001)
    b2[i] ~ dnorm(0, 0.0001)
  }
  
  for(i in 1:60){
    u0[i] ~ dnorm(0,1/vvint)
    u1[i] ~ dnorm(0, 1/vvslp)
    u2[i] ~ dnorm(0, 1/vvslp2)
  }
  
  vvslp2 ~ dgamma(1.1, 0.1)
  vvslp ~ dgamma(1.1, 0.1)
  vvint ~ dgamma(1.1, 0.1)
  vv ~ dgamma(1.1, 0.1)
}

"
newpig <- pig$newpig
trt = pig$trt
day = pig$day
day2 = pig$day2
weight = pig$weight

writeLines(mdl, 'fit11.txt')


data.jags <- c('weight', 'day', 'day2', 'trt', 'newpig')
parms <- c('b0' , 'b1', 'b2','u0','u1', 'vv', 'vvint', 'vvslp')

fit11 <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'fit11.txt', inits = NULL,
                  n.iter = 30000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
fit11

```

```{r, eval = FALSE}

sims <- as.mcmc(fit10)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
raftery.diag(sims)
effectiveSize(sims)


```


22.  Using DIC, which of the models that allow inference over all pigs would you choose?

Using DIC I would choose the model from question 20. This model just has random coefficients for the intercept and the linear day term. Adding a random effect for pig on the quadratic day term does not improve the model.

Problems 23 and 24 are not required.  If completed they will be worth bonus points.  10 points for problem 23, and 5 points
for problem 24.

23.  Since brm does not do well with these types of models, now you will do the work in Stan.  Create a hierarchical model with
hierarchical terms for both the intercepts and linear terms (slopes), but not for the quadratic term.  This is a hierarchical model
that is much like the random coefficients model in number 20.  Use normal priors with means of 0 and standard deviations of 100 where appropriate, and gamma(1.1,.5) priors for the variance terms.  What are means of the chains for the overall intercept in treatment 1, the overall
linear term coefficient (slope) in treatment 1, and the overall quadratic term coefficient in treatment 1?

Intercept for treatment 1 is 18.5, Linear term for treatment 1 is 0.85

```{r}

model <- "

data { 
  int <lower = 1> N;
  int q;
  real weight[N];
  int day[N];
  int trt[N];
}

parameters {
  real alpha[q];
  real beta[q];
  real mu_alpha;
  real mu_beta;
  real <lower = 0> serr;
  real <lower = 0> salpha;
  real <lower = 0> sbeta;
}


model {
  mu_alpha ~ normal(0, 100);
  mu_beta ~ normal(0, 100);
  alpha ~ normal(mu_alpha, salpha);
  beta ~ normal(mu_beta, sbeta);
  serr ~ gamma(1.1, 0.5);
  salpha ~ gamma(1.1, 0.5);
  sbeta ~ gamma(1.1, .5);
  for(i in 1:N){
    weight[i] ~ normal(alpha[trt[i]] + beta[trt[i]] * day[i], serr);
  }
}


generated quantities {
  vector[N] log_lik;
  real s2error;
  real s2int;
  real s2slope;
  s2error = serr*serr;
  s2int = salpha*salpha;
  s2slope = sbeta*sbeta;
  for (i in 1:N) log_lik[i] = normal_lpdf(weight[i] | alpha[trt[i]] + beta[trt[i]]*day[i], serr);
}

"

writeLines(model, 'hier.stan')
trt = pig$trt
day = pig$day
weight = pig$weight
N <- 360
q <- 3
hier_dat <- list(N=N, weight = weight, trt = trt, day = day, q=q)
fit12 <- stan(file = "hier.stan", data = hier_dat, iter = 11000, 
            warmup = 1000, chains = 4, thin = 2)


chains <- as.matrix(fit12)
dim(chains)
sims <- as.mcmc(chains)
```

```{r}
ggplot(data = pig, mapping  = aes(x = day, y = weight, col = trtf)) + 
  geom_point() + 
  geom_abline(intercept = 18.4, slope = 0.84, col = "orange") + 
  geom_abline(intercept = 19.38, slope = 0.82, col = "green") + 
  geom_abline(intercept = 17.664, slope = 0.663, col = "blue")
```


24. Compute the waic for the model in problem 23.

waic is 2145.3

```{r}
llfit <- extract_log_lik(fit12, parameter_name = "log_lik",
                         merge_chains = TRUE)
waic(llfit)
```

