---
title: "HW 3"
author: "Nathan Hawkins"
date: "2/5/2021"
output:
  word_document: default
  html_document: default
---

```{r}
twogroups <- read.table("C:/Users/nateh/Downloads/twogroups.dat")

names(twogroups) <- c("tmt", "y")
head(twogroups)
table(twogroups$tmt)
```

# 1 Jags Model

```{r}
library(R2jags)
hw3mdl <- "
  model {
  
    for(i in 1:33){
      y[i] ~ dnorm(mu[tmt[i]], 1/s2[tmt[i]])
    }
   
    #Priors
    for(i in 1:2){
      mu[i] ~ dnorm(125, .01)
      s2[i] ~ dunif(0, 2000)
    }
    
   
  }
"

writeLines(hw3mdl, 'hw3.txt')
y <- twogroups$y
tmt <- twogroups$tmt
data.jags <- c('y','tmt')
parms <- c("mu", "s2")

hw3.sim <- jags(data = data.jags, inits = NULL,
               parameters.to.save = parms,
               model.file = 'hw3.txt',
               n.iter = 16000,
               n.burnin = 1000,
               n.chains = 4,
               n.thin = 3)

hw3.sim


```

# Convergence checks
The plots look like they converge. Dependence is close to 1. Effective size is plenty high, and there's low autocorrelation.
```{r}
sims <- as.mcmc(hw3.sim)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)
head(sims)

plot(sims, type = 'l')

mu1 <- sims[,2]
mu2 <- sims[,3]
s2_1 <- sims[,4]
s2_2 <- sims[,5]

effectiveSize(sims)
autocorr.diag(sims)
raftery.diag(sims)
```


# Next Model
```{r}
library(R2jags)
a2mdl <- "
  model {
  
    for(i in 1:33){
      y[i] ~ dnorm(mu[tmt[i]], 1/s2[tmt[i]])
    }
   
    #Priors
    for(i in 1:2){
      mu[i] ~ dnorm(125, .01)
    }
    
    s2[1] ~ dgamma(2, .002)
    s2[2] ~ dgamma(2, .04)
    
   
  }
"

writeLines(a2mdl, 'a1.txt')
y <- twogroups$y
tmt <- twogroups$tmt
data.jags <- c('y','tmt')
parms <- c("mu", "s2")

a1.sim <- jags(data = data.jags, inits = NULL,
               parameters.to.save = parms,
               model.file = 'a1.txt',
               n.iter = 16000,
               n.burnin = 1000,
               n.chains = 4,
               n.thin = 3)

a1.sim

```
# Convergence checks

The plots look like they converge. Dependence is close to 1. Effective size is plenty high, and there's low autocorrelation.
```{r}
a1_sims <- as.mcmc(a1.sim)
a1_chains <- as.matrix(a1_sims)
a1_sims <- as.mcmc(a1_chains)
head(a1_sims)

plot(sims, type = 'l')

a1_mu1 <- a1_sims[,2]
a1_mu2 <- a1_sims[,3]
a1_s2_1 <- a1_sims[,4]
a1_s2_2 <- a1_sims[,5]

#Effective Size
effectiveSize(a1_sims)
#Autocorrelation
autocorr.diag(a1_sims)
#Diagnositcs
raftery.diag(a1_sims)
```



# Comparing Models
DIC is lower for the model with the gamma variances.
Here are density plots of the differences

```{r}
diff1 <- mu1-mu2
diff2 <- a1_mu1 - a1_mu2
plot(density(diff1), col = 'blue', lwd = 4, main = "Comparison of the distribution of the difference of mu's")
lines(density(diff2), col = 'orchid', lwd = 4)
legend(-10, .03, c("Uniform Prior variance", "Gamma Prior Varaince"), col = c("blue", "orchid"), lty = 1, lwd = 4)

```
# 2 Same thing but in sas
Sas Code 



 library(SASmarkdown)

saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"
knitr::opts_chunk$set(engine="sas", engine.path=saspath,
        engine.opts=sasopts, comment=NA)

knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts


# Sas Code. Works in R just not when I knit it.


 data twogroups;
 infile 'C:/Users/nateh/Downloads/twogroups.dat';
 input tmt y;
 run;

 proc means data=twogroups;
 run;



proc mcmc data = twogroups nbi = 30000 nmc = 300000 thin = 30 outpost = 'C:/Users/nateh/Documents/Stat 451/anovamodel_1.sas7bdat' dic propcov=quanew            monitor=(_parms_) stats = all diagnostics = all;
array mu[2] mu1-mu2;
array vv[2] vv1-vv2;
parms mu1: 0;
parms mu2: 0;
parms vv1: 10;
parms vv2: 10;
prior mu1: ~ normal(125, var = 100);
prior mu2: ~ normal(125, var = 100);
prior vv1: ~ uniform(0, 2000);
prior vv2: ~ uniform(0, 2000);
model y ~ normal(mu[tmt], var = vv[tmt]);
run;

data twogroups;
infile 'C:/Users/nateh/Downloads/twogroups.dat';
input tmt y;
run;

proc means data=twogroups;
run;

 proc mcmc data = twogroups nbi = 30000 nmc = 300000 thin = 30 outpost = 'C:/Users/nateh/Documents/Stat 451/anovamodel_2.sas7bdat' dic propcov=quanew            monitor=(_parms_) stats = all diagnostics = all;
array mu[2] mu1-mu2;
array vv[2] vv1-vv2;
parms mu1: 0;
parms mu2: 0;
parms vv1: 10;
parms vv2: 10;
prior mu1: ~ normal(125, var = 100);
prior mu2: ~ normal(125, var = 100);
prior vv1: ~ gamma(2, scale = 500);
prior vv2: ~ gamma(2, scale = 25);
model y ~ normal(mu[tmt], var = vv[tmt]);
run;
endsas;





#Check Convergence for sas model
```{r echo = FALSE}

library(haven)
library(coda)

anovamodel1 <- read_sas("C:/Users/nateh/Documents/Stat 451/anovamodel_1.sas7bdat")
anovamodel2 <- read_sas("C:/Users/nateh/Documents/Stat 451/anovamodel_2.sas7bdat")


plot(anovamodel1$mu1, type = 'l')
plot(anovamodel1$mu2, type = 'l')
plot(anovamodel1$vv1, type = 'l')
plot(anovamodel1$vv2, type = 'l')

an_1.sims <- as.mcmc(anovamodel1)
an_1.sims.chains <- as.matrix(an_1.sims)
an_1.sims <- as.mcmc(an_1.sims.chains)
head(an_1.sims)

#Effective Size
effectiveSize(an_1.sims)
#Autocorrelation
autocorr.diag(an_1.sims)
#Diagnositcs
raftery.diag(an_1.sims)

an_2.sims <- as.mcmc(anovamodel2)
an_2.sims.chains <- as.matrix(an_2.sims)
an_2.sims <- as.mcmc(an_2.sims.chains)
head(an_2.sims)

effectiveSize(an_2.sims)
autocorr.diag(an_2.sims)
raftery.diag(an_2.sims)

```

```{r}
plot(density(anovamodel2$mu1 - anovamodel2$mu2), col = 'blue', lwd = 2, main = "Mu1-Mu2 for both models compared")
lines(density(anovamodel1$mu1 - anovamodel2$mu2), col = 'red', lwd = 2)
legend(-10, .03, c("Prior Variance follows gamma", "Prior Variance follows uniform"), col = c('blue', 'red'), lty = 1,lwd = 2)

```














# 3 Prior Predictive for uniform
Prior Predictive Distributions
```{r}
# Generate prior distribution
prpred_1 <- NULL
for(i in 1:1000){
  prpred_1[i] <- rnorm(1, 125, sqrt(100)) + rnorm(1,0,sqrt(runif(1,0,2000)))
}
#Plot Prior
plot(density(prpred_1), type = 'l', main = "Prior Distribution")

#Plot Prior with treatments
hist(twogroups[twogroups$tmt == 1,]$y, freq =  FALSE, main = "Treatment 1 histogram with prior distribution", xlim = c(30, 250), ylim = c(0, 0.06))
lines(density(prpred_1), type = 'l', col = 'blue', lwd = 2)
hist(twogroups[twogroups$tmt == 2,]$y, freq =  FALSE, main = "Treatment 2 histogram with prior distribution", xlim = c(30, 250))
lines(density(prpred_1), type = 'l', col = 'blue', lwd = 2)
```


# 4 Prior Predictive for gamma model
Prior Predictive Distributions
```{r}
# Generate prior distribution
prpred_2 <- NULL
for(i in 1:1000){
  prpred_2[i] <- rnorm(1, 125, sqrt(100)) + rnorm(1,0,sqrt(rgamma(1, 2, 0.002)))
}

prpred_3 <- NULL
for(i in 1:1000){
  prpred_3[i] <- rnorm(1, 125, sqrt(100)) + rnorm(1,0,sqrt(rgamma(1, 2, 0.04)))
}


#Plot Prior
plot(density(prpred_2), type = 'l', main = "Prior Distribution")
plot(density(prpred_3), type = 'l', main = "Prior Distribution")

#Plot Prior with treatments
hist(twogroups[twogroups$tmt == 1,]$y, freq =  FALSE, main = "Treatment 1 histogram with gamma prior distribution", xlim = c(30, 250), ylim = c(0, 0.06))
lines(density(prpred_2), type = 'l', col = 'blue', lwd = 2)
hist(twogroups[twogroups$tmt == 2,]$y, freq =  FALSE, main = "Treatment 2 histogram with gamma prior distribution", xlim = c(90, 150))
lines(density(prpred_3), type = 'l', col = 'blue', lwd = 2)
```

# 5 Posterior Predictive from Jags model
Posterior Predictive Distributions
```{r}
#Posterior Predictive
popdat1 <- NULL
for(i in 1:20000){
            #Noise in mean, #Noise in Error
  popdat1[i] <- mu1[i] + rnorm(1,0,sqrt(s2_1[i]))
}

hist(twogroups$y[tmt == 1], breaks = 5, freq = FALSE, xlim = c(70, 200), main = "Histogram of Treatment 1 with Posterior Predictive line", ylim = c(0, 0.06))
lines(density(popdat1))
hist(twogroups$y[tmt == 2], breaks = 5, freq = FALSE, xlim = c(70, 200), main = "Histogram of Treatment 2 with Posterior Predictive line")
lines(density(popdat1))

```

# 6 Posterior Predictive for SAS model
Posterior Predictive Distributions for the 2nd model. These posterior predictive lines closely follow the histogram
```{r}
#Posterior Predictive
sas_mu1 <- an_2.sims[,2]
sas_mu2 <- an_2.sims[,3]
sas_vv1 <- an_2.sims[,4]
sas_vv2 <- an_2.sims[,5]

popdat2_sas <- NULL

for(i in 1:10000){
            #Noise in mean, #Noise in Error
  popdat2_sas[i] <- sas_mu1[i] + rnorm(1,0,sqrt(sas_vv1[i]))
}

popdat3_sas <- NULL

for(i in 1:10000){
  popdat3_sas[i] <- sas_mu2[i] + rnorm(1,0,sqrt(sas_vv2[i]))
}

hist(twogroups$y[tmt == 1], breaks = 5, freq = FALSE, xlim = c(70, 200), main = "Histogram of Treatment 1 with Posterior Predictive line", ylim = c(0, 0.06))
lines(density(popdat2_sas))
hist(twogroups$y[tmt == 2], breaks = 5, freq = FALSE, xlim = c(70, 200), main = "Histogram of Treatment 2 with Posterior Predictive line")
lines(density(popdat3_sas))
```



