---
title: "HW 4"
author: "Nathan Hawkins"
date: "2/11/2021"
output:
  word_document: default
  html_document: default
---

```{r}
vo2 <- read.table("vo2.dat", header = TRUE)
head(vo2)
```

# 1 
Running MLR with the 6 covariates, I get a DIC of 756.3.

```{r}
library(R2jags)
mdl <- "
  model {
  
  for(i in 1:120){
    y[i] ~ dnorm(mu[i], prec)
    mu[i] <- b0 +  bgen*gender[i] + bbmi*bmi[i] + brpe*rpe[i]+ bhr*hr[i] + bmph*mph[i] + bage*age[i]
  }
  
  b0 ~ dnorm(0 , 0.001)
  bgen ~ dnorm(0, 0.001)
  bbmi ~ dnorm(0, 0.001)
  brpe ~ dnorm(0, 0.001)
  bhr ~ dnorm(0, 0.001)
  bmph ~ dnorm(0, 0.001)
  bage ~ dnorm(0, 0.001)
  
  prec <- 1/vv
  vv ~ dgamma(1.5, 0.2)
  
  }

"
#Assing simple names for covariates
mph <- vo2$MPH
bmi <- vo2$BMI
gender <- vo2$Gender
y <- vo2$MaxVO2ML
rpe <- vo2$RPE
hr <- vo2$HR
age <- vo2$Age1


# Standardize the covariates
mph <- (mph-mean(mph))/sd(mph)
bmi <- (bmi-mean(bmi))/sd(bmi)
gender <- (gender-mean(gender))/sd(gender)
rpe <- (rpe-mean(rpe))/sd(rpe)
age <- (age-mean(age))/sd(age)
hr <- (hr-mean(hr))/sd(hr)
mphhr <- mph*hr





writeLines(mdl, 'vo2.txt')

#What to use
data.jags <- c('y', 'gender', 'bmi', 'rpe', 'mph', 'hr', 'age')

#What to call it in the output
parms.jags <- c('b0', "brpe", 'bgen', 'bbmi', 'bmph', 'vv', 'bhr', 'bage')

vo2.sim <- jags(data= data.jags, parameters.to.save = parms.jags,
                  model.file = 'vo2.txt', inits = NULL,
                  n.iter = 6000, n.thin = 1, n.chains = 4,
                  n.burnin = 1000)


vo2.sim
```

# 2
Running the same model but without age, I get a DIC of 755.1.

```{r}
library(R2jags)
mdl2 <- "
  model {
  
  for(i in 1:120){
    y[i] ~ dnorm(mu[i], prec)
    mu[i] <- b0 +  bgen*gender[i] + bbmi*bmi[i] + brpe*rpe[i]+ bhr*hr[i] + bmph*mph[i]
  }
  
  b0 ~ dnorm(0 , 0.001)
  bgen ~ dnorm(0, 0.001)
  bbmi ~ dnorm(0, 0.001)
  brpe ~ dnorm(0, 0.001)
  bhr ~ dnorm(0, 0.001)
  bmph ~ dnorm(0, 0.001)
  
  prec <- 1/vv
  vv ~ dgamma(1.5, 0.2)
  
  }

"
#Assing simple names for covariates
mph <- vo2$MPH
bmi <- vo2$BMI
gender <- vo2$Gender
y <- vo2$MaxVO2ML
rpe <- vo2$RPE
hr <- vo2$HR
age <- vo2$Age1


# Standardize the covariates
mph <- (mph-mean(mph))/sd(mph)
bmi <- (bmi-mean(bmi))/sd(bmi)
gender <- (gender-mean(gender))/sd(gender)
rpe <- (rpe-mean(rpe))/sd(rpe)
age <- (age-mean(age))/sd(age)
hr <- (hr-mean(hr))/sd(hr)






writeLines(mdl2, 'vo2_v2.txt')

#What to use
data.jags <- c('y', 'gender', 'bmi', 'rpe', 'mph', 'hr')

#What to call it in the output
parms.jags <- c('b0', "brpe", 'bgen', 'bbmi', 'bmph', 'vv', 'bhr')

vo2_v2.sim <- jags(data= data.jags, parameters.to.save = parms.jags,
                  model.file = 'vo2_v2.txt', inits = NULL,
                  n.iter = 6000, n.thin = 1, n.chains = 4,
                  n.burnin = 1000)


vo2_v2.sim
```


# 3 
I would choose the second model because it has a lower DIC so it is better at predicting with fewer covariates.


# 4
Same model but with mph*hr. This model has a DIC of 753.5.

```{r}
library(R2jags)
mdl3 <- "
  model {
  
  for(i in 1:120){
    y[i] ~ dnorm(mu[i], prec)
    mu[i] <- b0 +  bgen*gender[i] + bbmi*bmi[i] + brpe*rpe[i]+ bhr*hr[i] + bmph*mph[i] + bmphhr*mphhr[i]
  }
  
  b0 ~ dnorm(0 , 0.001)
  bgen ~ dnorm(0, 0.001)
  bbmi ~ dnorm(0, 0.001)
  brpe ~ dnorm(0, 0.001)
  bhr ~ dnorm(0, 0.001)
  bmph ~ dnorm(0, 0.001)
  bmphhr ~ dnorm(0, .001)
  
  prec <- 1/vv
  vv ~ dgamma(1.5, 0.2)
  
  }

"
#Assing simple names for covariates
mph <- vo2$MPH
bmi <- vo2$BMI
gender <- vo2$Gender
y <- vo2$MaxVO2ML
rpe <- vo2$RPE
hr <- vo2$HR
age <- vo2$Age1


# Standardize the covariates
mph <- (mph-mean(mph))/sd(mph)
bmi <- (bmi-mean(bmi))/sd(bmi)
gender <- (gender-mean(gender))/sd(gender)
rpe <- (rpe-mean(rpe))/sd(rpe)
age <- (age-mean(age))/sd(age)
hr <- (hr-mean(hr))/sd(hr)
mphhr <- mph*hr
mphhr = (mphhr - mean(mphhr))/sd(mphhr)





writeLines(mdl3, 'vo2_v3.txt')

#What to use
data.jags <- c('y', 'gender', 'bmi', 'rpe', 'mph', 'hr', 'mphhr')

#What to call it in the output
parms.jags <- c('b0', "brpe", 'bgen', 'bbmi', 'bmph', 'vv', 'bhr', 'bmphhr')

vo2_v3.sim <- jags(data= data.jags, parameters.to.save = parms.jags,
                  model.file = 'vo2_v3.txt', inits = NULL,
                  n.iter = 6000, n.thin = 1, n.chains = 4,
                  n.burnin = 1000)


vo2_v3.sim
```


# 5
I prefer the last model because it has the lowest DIC. Adding the interaction term is worth it.


# 6 SAS code works, doesn't knit.


 library(SASmarkdown)

saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"
knitr::opts_chunk$set(engine="sas", engine.path=saspath,
        engine.opts=sasopts, comment=NA)

knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts




# 1 Sas
DIC is 756.5




data vo2;
	infile 'C:/Users/nateh/Documents/Stat 451/vo2.dat';
 	input ID Gender Age BMI MPH HR RPE MAXVO2ML;
run;

proc standard data=vo2 out=vo2 mean=0 std=1;
var ID Gender Age BMI MPH HR RPE:;
run;

 
proc mcmc data = vo2 
	nbi = 30000 nmc = 300000 thin = 30 
	outpost = 'C:/Users/nateh/Documents/Stat 451/MLRHW4.sas7bdat' dic propcov=quanew  
    monitor=(_parms_) stats = all;
parms b0 0;
parms bmph 0;
parms bage 0;
parms bhr 0;
parms brpe 0;
parms bgen 0;
parms bbmi 0;
parms vv 1.5;
prior b0 ~ normal(0, var = 1000);
prior bmph: ~ normal(0, var = 1000);
prior bage: ~ normal(0, var = 1000);
prior bhr: ~ normal(0, var = 1000);
prior brpe: ~ normal(0, var = 1000);
prior bbmi: ~ normal(0, var = 1000);
prior bgen: ~ normal(0, var = 1000);
prior vv ~ gamma(1.5, scale = 5);
mu = b0 +  bgen*Gender + bbmi*BMI + brpe*RPE + bhr*HR + bmph*MPH + bage*Age;
model MAXVO2ML ~ normal(mu, var = vv);
run;


# 2 
Model without age, DIC is 755.1


data vo2;
	infile 'C:/Users/nateh/Documents/Stat 451/vo2.dat';
 	input ID Gender Age BMI MPH HR RPE MAXVO2ML;
run;

proc standard data=vo2 out=vo2 mean=0 std=1;
var ID Gender Age BMI MPH HR RPE:;
run;

 
proc mcmc data = vo2 
	nbi = 30000 nmc = 300000 thin = 30 
	outpost = 'C:/Users/nateh/Documents/Stat 451/MLRHW4_2.sas7bdat' dic propcov=quanew  
    monitor=(_parms_);
parms b0 0;
parms bmph 0;
parms bhr 0;
parms brpe 0;
parms bgen 0;
parms bbmi 0;
parms vv 1.5;
prior b0 ~ normal(0, var = 1000);
prior bmph: ~ normal(0, var = 1000);
prior bhr: ~ normal(0, var = 1000);
prior brpe: ~ normal(0, var = 1000);
prior bbmi: ~ normal(0, var = 1000);
prior bgen: ~ normal(0, var = 1000);
prior vv ~ gamma(1.5, scale = 5);
mu = b0 +  bgen*Gender + bbmi*BMI + brpe*RPE + bhr*HR + bmph*MPH;
model MAXVO2ML ~ normal(mu, var = vv);
run;


# 3 Sas 
2nd model is better, lower DIC

# 4 SAS
Model without age, with mphhr. DIC is 753.6

data vo2;
	infile 'C:/Users/nateh/Documents/Stat 451/vo2.dat';
 	input ID Gender Age BMI MPH HR RPE MAXVO2ML;
run;

proc standard data=vo2 out=vo2 mean=0 std=1;
var ID Gender Age BMI MPH HR RPE:;
run;

 
proc mcmc data = vo2 
	nbi = 30000 nmc = 300000 thin = 30 
	outpost = 'C:/Users/nateh/Documents/Stat 451/MLRHW4_3.sas7bdat' dic propcov=quanew  
    monitor=(_parms_);
parms b0 0;
parms bmph 0;
parms bhr 0;
parms brpe 0;
parms bgen 0;
parms bbmi 0;
parms binteraction 0;
parms vv 1.5;
prior b0 ~ normal(0, var = 1000);
prior bmph: ~ normal(0, var = 1000);
prior bhr: ~ normal(0, var = 1000);
prior brpe: ~ normal(0, var = 1000);
prior bbmi: ~ normal(0, var = 1000);
prior bgen: ~ normal(0, var = 1000);
prior binteraction ~ normal(0, var = 1000);
prior vv ~ gamma(1.5, scale = 5);
mu = b0 +  bgen*Gender + bbmi*BMI + brpe*RPE + bhr*HR + bmph*MPH + binteraction*(MPH*HR);
model MAXVO2ML ~ normal(mu, var = vv);
run;


# 5 SAS
The results are the same, the third model is the best.