---
title: "HW 7"
author: "Nathan Hawkins"
date: "3/11/2021"
output:
  word_document: default
  html_document: default
---

```{r}
library(ggplot2)
library(dplyr)
library(brms)
library(R2jags)
pbib = read.table("pbib.dat", header = TRUE)
head(pbib)

pbib$trtf <- as.factor(pbib$trt)
ggplot(data = pbib, mapping = aes(x = trtf, y = response, col = trtf)) + 
  geom_boxplot()

range(pbib$blk)
range(pbib$trt)
dim(pbib)
```
If we want to make inference on treatment, then we need to account for block to block variability.
# 1. Mixed Model
The effects for s2error and s2block are .098 and .075, respectively.
```{r}
library(R2jags)
mdl <- "

model {

  for(i in 1:60){
    response[i] ~ dnorm(mu[i], 1/s2error)
    mu[i] <- beta[trt[i]] + u[blk[i]]
  }
  
  # Priors
  for(i in 1:15){
    beta[i] ~ dnorm(0, 0.01)
  }
  

  for(i in 1:15){
    u[i] ~ dnorm(0, 1/s2blk)
  }
  
  s2error ~ dgamma(1.1, 0.1)
  s2blk ~ dgamma(1.1, 0.1)

}
"

writeLines(mdl, 'dep.txt')

response <- pbib$response
trt <- pbib$trt
blk <- pbib$blk

data.jags <- c('response', 'trt', 'blk')
parms <- c('beta','s2error', 's2blk')

mixedmods.sim <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'dep.txt', inits = NULL,
                  n.iter = 12000, n.thin = 5, n.burnin = 2000,
                  n.chains = 5)
mixedmods.sim

```

Raftery Lewis Diagnostics look good; they are around 0. So do effective sample sizes; they are all over 5,000.
```{r}
sims <- as.mcmc(mixedmods.sim)
chains <- as.matrix(sims)
raftery.diag(chains)
effectiveSize(chains)
colnames(chains)


```

# 2)
The 5 strains with the highest responses are 12, 6, 9, 11, 15 

# 3)
The top five are not statistically significant from each other.
But there is a significant difference between strain 12 and strain 2.
```{r}
quantile(chains[,8] - chains[,3], probs = c(.025, .975))
quantile(chains[,9] - chains[,3], probs = c(.025, .975))

```

# 4) Now do it all in brm
Effective sample sizes look good.
The top 5 strains are 12, 6, 9, 11, 15. These are the same as before
```{r}
library(brms)

fit2 <- brm(formula = response ~ -1 + trtf + (1|blk), data = pbib, 
            family = "gaussian",
            prior = c(set_prior("normal(0,10)", class = "b"),
                      set_prior("gamma(1.1,.1)", class = "sigma"),
                      set_prior("gamma(1.1,.1", class = 'sd')),
            warmup = 1000, iter = 3000, chains = 4, 
            control = list(adapt_delta = 0.98),
            save_pars = save_pars(all = TRUE))
summary(fit2)
```
p_loo is 23.1 for this analysis.
```{r}
library(loo)
loo1 <- loo(fit2)
```

