---
title: "HW 8"
author: "Stat 451"
date: "3/22/2021"
output:
  word_document: default
  pdf_document: default
---

For this assignment you will use the data file 'outside.dat'.  In this file there
are two columns, time and response.  The response is a binary variable indicating
whether or not a set in volleyball resulted in a kill.  The time is the elapsed
time in seconds the ball was in the air after leaving the setter's hands until 
being struck by the outside hitter.  The question is whether there is a 
relationship between set speed (i.e. time the ball was in the air) and probability
of the set resulting in a kill.  The problem should be done as a linear logistic
regression.

Likelihood will be bernouli
```{r}
outside <- read.table("outside.dat", header = TRUE)
dim(outside)
head(outside)
library(R2jags)
library(ggplot2)

ggplot(data = outside, mapping = aes(x= time, y = as.factor(response))) + 
  geom_boxplot()
```


1.  Write the code to analyze the data in JAGS.

```{r}

mdl <- "
  model {
    
    for(i in 1:718){
      response[i] ~ dbern(p[i])
      logit(p[i]) <- b0 + b1*time[i]
    }
     # Prior Values
      b1 ~ dnorm(0, 0.001)
      b0 ~ dnorm(0, 0.001)
  }
"

# Reponse is binomial
# The effects are on the logit scale

writeLines(mdl, 'logistic.txt')
response <- outside$response
time <- outside$time

data.jags <- c('time', "response")
parms <- c('b0','b1')
logistic.sim <- jags(data = data.jags, inits = NULL, parameters.to.save = parms,
                     model.file = 'logistic.txt', n.iter = 12000, n.burnin = 2000,
                     n.chains = 8, n.thin = 10)

logistic.sim

```





2. Show conversion diagnostics.  At the very least the Raftery-Lewis and the
effective sample size.

Diagnostics are good.
```{r}
sims <- as.mcmc(logistic.sim)
chains <- as.matrix(sims)
raftery.diag(chains)
effectiveSize(chains)
```


3. Is there a significant relationship between time and the probability of 
a kill?  How do you know?  What are the $\hat\beta$'s?

Since the 95% posterior credible interval does not span 0 then there is a significant effect.
Yes there is a significant relationship between time and probability of a kill. A density plot of the changes in probability associated with a 1 second time change does not contain 0. So I conclude that it is significant.
```{r}
exp(-1.665)/(1+exp(-1.665))
exp.b1s <- exp(chains[,2])/(1+exp(chains[,2]))

plot(density(exp.b1s))
colnames(chains)

plot(density(chains[,2]))
```


4.  Produce a plot with probability of a kill on the y-axis and time
on the x-axis.  Show the best fit line.

```{r}
min(outside$time)
max(outside$time)

plot(outside$time, outside$time*-0.159)

p1 <- exp(chains[,1] + chains[,2])/(exp(1+chains[,1]+chains[,2]))
p1 <- chains[,1] + chains[,2]
prob1 <- exp(p1)/(1+exp(p1))
quantile(p1,c(0.25,0.975))

length(prob1)
tt <- seq(0.025, 2.1, by = .1)
length(tt)
probs <- matrix(0,8000,21)
for(i in 1:21){
  probs[,i] <- exp(chains[,1]+chains[,2]* tt[i])/(1+exp(chains[,1]+chains[,2]*tt[i]))
}

length(tt)
plot(tt, apply(probs,2,mean), type = 'l')

```


5.  Add the 95\% posterior probability intervals to the plot in the
previous problem.

```{r}
exp.b1s <- exp(chains[,2])/(1+exp(chains[,2]))
min(outside$time)
max(outside$time)

plot(outside$time, outside$time*-0.159)

p1 <- exp(chains[,1] + chains[,2])/(exp(1+chains[,1]+chains[,2]))
p1 <- chains[,1] + chains[,2]
prob1 <- exp(p1)/(1+exp(p1))
quantile(p1,c(0.25,0.975))

length(prob1)
tt <- seq(0.025, 2.1, by = .1)
length(tt)
probs <- matrix(0,8000,21)
for(i in 1:21){
  probs[,i] <- exp(chains[,1]+chains[,2]* tt[i])/(1+exp(chains[,1]+chains[,2]*tt[i]))
}

length(tt)
plot(tt, apply(probs,2,mean), type = 'l')
lines(tt,apply(probs,2,quantile,0.025), col = "blue")
lines(tt,apply(probs,2,quantile,0.975), col = "blue")
```


6.  Now write the code to analyze the problem in Stan.  Compare
the $\beta$'s you got in JAGS to the $\beta$'s you got in Stan.
Provide plots of the posteriors for the $\beta_0$'s on one
set of axes.  Provide plots of the posteriors for the $\beta_1$'s on one
set of axes.

The effects for intercept and slope are almost the same as the model in jags. 

```{r}
head(outside)
dim(outside)

N <- nrow(outside)
x <- outside$time
y <- outside$response
out_dat <- list(N=N, y = y, x = x)
library(rstan)
logistic_stan.sim <- stan(file = "outside.stan", data = out_dat, iter = 5000, 
              warmup = 1000, chains = 1, thin = 2)

logistic_stan.sim
```

Density Plots

```{r}
sims2 <- as.matrix(logistic_stan.sim)
chains2 <- as.mcmc(sims2)


plot(density(chains[,2]), main = "Density plot of B1 posterior draws", lwd = 2)
lines(density(chains2[,2]), col = "blue", lwd = 2)

plot(density(chains[,1]), main = "Density plot of B0 posterior draws", lwd = 2)
lines(density(chains2[,1]), col = "blue", lwd = 2)
```

