---
title: "HW 5"
author: "Nathan Hawkins"
date: "2/26/2021"
output:
  word_document: default
  html_document: default
---

```{r}
crack <- read.table("C:/Users/nateh/Downloads/cracker.dat", header = TRUE)

```
# 1 Different slopes and intercepts for each treatment

```{r}
library(R2jags)
head(crack)



mdl <- 
  "model {
  
  for(i in 1:14){
    sales[i] ~ dnorm(mu[i], 1/vv);
    mu[i] <- b0[tmt[i]] + b1[tmt[i]]*presales[i];
  }
  
  for(i in 1:3){
  b0[i] ~ dnorm(25, .001);
  b1[i] ~ dnorm(0, .001);
  }
  
  vv ~ dgamma(1.5, .5);
  
  }
  

"

writeLines(mdl, 'ancova.txt')
sales <- crack$sales
presales <- crack$presales
tmt <- crack$tmt

data.jags <- c('sales', 'presales', 'tmt')
parms <- c('b0', 'b1', 'vv')

ancova.sim <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'ancova.txt', inits = NULL,
                  n.iter = 500000, n.thin = 500, n.burnin = 1000,
                  n.chains = 5)
ancova.sim  
```
# 2 
Check the differences between the slopes. There doesn't appear to be any significant difference in slope.

```{r}
sims <- as.mcmc(ancova.sim)
chains <- as.matrix(sims)
sims <- as.mcmc(chains)


b11 <- chains[,4]
b12 <- chains[,5]
b13 <- chains[,6]

plot(density(b11 - b12))
plot(density(b11 - b13))
plot(density(b13 - b12))

quantile(b11-b12, probs = c(.025, .975))
quantile(b11-b13, probs = c(.025, .975))
quantile(b13-b12, probs = c(.025, .975))


```

# 3 Same slope for each treatment. Different Intercepts

```{r}
library(R2jags)
head(crack)



mdl2 <- 
  "model {
  
  for(i in 1:14){
    sales[i] ~ dnorm(mu[i], 1/vv);
    mu[i] <- b0[tmt[i]] + b1*presales[i];
  }
  
  for(i in 1:3){
    b0[i] ~ dnorm(3, .001);
  }
  
  b1 ~ dnorm(0, .01);
  vv ~ dgamma(1, .125);
  
  }
  

"

writeLines(mdl2, 'ancova2.txt')
sales <- crack$sales
presales <- crack$presales
tmt <- crack$tmt

data.jags <- c('sales', 'presales', 'tmt')
parms <- c('b0', 'b1', 'vv')

ancova2.sim <- jags(data= data.jags, parameters.to.save = parms,
                  model.file = 'ancova2.txt', inits = NULL,
                  n.iter = 100000, n.thin = 2, n.burnin = 1000,
                  n.chains = 5)
ancova2.sim  

```

This second model gives a lower DIC (65.1) than the first model (66.2). I believe that one slope should be enough.


# 4

Using 28 as a comparison point. Plots of differences in the values at presales = 28. Both the plots and the quantile show that there is a signifanct difference between treatments one and three, and treatments two and three. Treatments one and two, however, are not statistically significant.

# SAS Code

data cracker;
infile 'C:/Users/nateh/Downloads/cracker.dat';
input tmt presales sales;
run;



proc mcmc data = cracker outpost = 'C:/Users/nateh/Documents/Stat 451/ancova_sas.sas7bdat' seed = 1234 
nmc = 5000000 nbi= 1000 monitor=(_parms_) thin = 200 diagnostics=(rl ess autocorr) dic propcov=quanew;
  array b0[3] b01-b03;
  parms b01 0;
  parms b02 0;
  parms b03 0; 
  parms b11 0;
  parms vv 20;
  prior b01: ~ normal(0, var = 1000);
  prior b02: ~ normal(0, var = 1000);
  prior b03: ~ normal(0, var = 1000);
  prior b11: ~ normal(0, var = 1000);
  prior vv ~ gamma(1, scale = 8);
  mu = b0[tmt]+b11*28;
  model sales ~ normal(mu, var = vv);
run;

  

# R code

```{r}
library(haven)
sasoutput <- read_sas('ancova_sas.sas7bdat')
sas.sims <- as.mcmc(sasoutput)
sas.chains <- as.matrix(sas.sims)

raftery.diag(sas.chains)
colnames(sas.chains)

sasb01 = sas.chains[,2]
sasb02 = sas.chains[,3]
sasb03 = sas.chains[,4]

sasb1 = sas.chains[,5]


mean(sasb01)
mean(sasb02)
mean(sasb03)

mean(sasb1)

newchain1 <- sasb1*28 + sasb01
newchain2 <- sasb1*28 + sasb02
newchain3 <- sasb1*28 + sasb03

mean(newchain1)
mean(newchain2)
mean(newchain3)


plot(density(newchain1 - newchain2))
plot(density(newchain1 - newchain3))
plot(density(newchain2 - newchain3))

quantile(newchain1 - newchain2, probs = c(.025, .975))
quantile(newchain1 - newchain3, probs = c(.025, .975))
quantile(newchain2 - newchain3, probs = c(.025, .975))

```

