---
title: "Proc MCMC"
author: "Nathan Hawkins"
date: "2/2/2021"
output:
  word_document: default
  html_document: default
---



```{r}
library(haven)
```

```{r setup}
library(SASmarkdown)

# set up the options so that knit knows where you SAS executable is
# set the linesize to be easily readable on letter size paper, portrait
# and set the knir options using opts_chunk$set().

saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"
knitr::opts_chunk$set(engine="sas", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)

knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

```{sas ex-sas}

data westwood;
infile 'C:/Users/nateh/Documents/Stat 451/westwood.dat';
input x y;
run;

proc means data=westwood;
run;



proc mcmc data = westwood outpost = 'C:/Users/nateh/Documents/Stat 451/linregpost.sas7bdat' seed = 1234 nmc = 100000 nbi= 10000 monitor=(_parms_) thin = 10 diagnostics=(rl ess autocorr) dic propcov=quanew;
  parms b0 0 b1 0;
  parms s2 20;
  prior b0 ~ normal(0, var = 10000);
  prior b1 ~ normal(0, var = 10000);
  prior s2 ~ gamma(1.5, scale = 2);
  mu = b0+b1*x;
  model y ~ normal(mu, var = s2);
  run;
  endsas;

```


# westwoodmcmc <- read_sas('C:/Users/nateh/Documents/Stat 451/linregpost.sas7bdat')
# dim(westwoodmcmc)
# plot(westwoodmcmc$b0, type  = 'l')
# plot(westwoodmcmc$b1, type = 'l')
# plot(westwoodmcmc$s2, type = 'l')


# Anova in Sas
```{sas}

data westwood;
infile 'C:/Users/nateh/Documents/Stat 451/westwood.dat';
input x y;
run;

proc means data=westwood;
run;



proc mcmc data = westwood outpost = 'C:/Users/nateh/Documents/Stat 451/linregpost.sas7bdat' seed = 1234 nmc = 100000 nbi= 10000 monitor=(_parms_) thin = 10 diagnostics=(rl ess autocorr) dic propcov=quanew;
  parms b0 0 b1 0;
  parms s2 20;
  prior b0 ~ normal(0, var = 10000);
  prior b1 ~ normal(0, var = 10000);
  prior s2 ~ gamma(1.5, scale = 2);
  mu = b0+b1*x;
  model y ~ normal(mu, var = s2);
  run;
  endsas;


```

```{sas}
data anova;
infile 'C:/Users/nateh/Downloads/anova.dat';
input x y;
run;

proc means data=anova;
run;
```


```{sas ex-sas}
data anova;
infile 'C:/Users/nateh/Downloads/anova (1).dat';
input tmt rsp dummy;
run;

proc means data=anova;
run;
```

proc mcmc data = anova nbi = 30000 nmc = 300000 thin = 30 outpost = 'C:/Users/nateh/Documents/Stat 451/anovaout.sas7bdat' dic propcov=quanew            monitor=(_parms_) stats = all diagnostics = all;
array mu[4] mu1-mu4;
array vv[4] vv1-vv4;
parms mu: 0;
parms vv: 10;
prior mu: ~ normal(10, var = 1000);
prior vv: ~ gamma(1.5, scale = 10);
model rsp ~ normal(mu[tmt], var = vv[tmt]);
run;
endsas;


```
# Do prior vv[1] ~ gamma(whatever)
